---
title: "Standardization of Laboratory Data using zlog"
author: "Sandra Klawitter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardization of Laboratory Data using zlog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
The zlog() function and its related utilities provide methods for standardizing and normalizing laboratory measurement data using reference limits. The lower reference limits (LL) and upper reference limits (UL) can transform any result x into a zlog value using the following equation: zlog(x) = (log(x)–(log(LL)+ log(UL))/2)*3.92/(log(UL)–log(LL)). Values ranging from –1.96 to 1.96 are considered normal, while values below –5 and above 5 indicate pathological conditions.

## Usage
Below is a basic demonstration of applying zlog transformations.

```{r setup, message = FALSE}
library(znorm)
```

We can transform laboratory measurements based on defined lower and upper reference limits:

```{r example}
# Example analyte measurements
vals <- c(10, 15, 20, 25, 30, 100)

# Calculate zlog values using Lower limit (L) and Upper limit (U)
zlog_scores <- zlog(vals, limits = c(12, 25))

# Print standardized zlog scores
print(zlog_scores)
```

Use the same reference limits to recover the original measurements:

```{r inverse}
izlog(zlog_scores, limits = c(12, 25))
```

### References

Hoffmann, Georg & Klawonn, Frank & Lichtinghagen, Ralf & Orth, Matthias. (2018). The zlog value as a basis for the standardization of laboratory results. LaboratoriumsMedizin. 41. 10.1515/labmed-2017-0135. 
