---
title: "Standardization of Laboratory Data using znorm"
author: "Sandra Klawitter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standardization of Laboratory Data using znorm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
The znorm() function and its related utilities provide methods for Box-Cox standardizing and normalizing laboratory measurement data using reference limits. The lower reference limits (LL) and upper reference limits (UL) can transform any result x into a znorm value using the following equation. Values ranging from –1.96 to 1.96 are considered normal, while values below –5 and above 5 indicate pathological conditions. For lambda = 0 the result equals of zlog() for lambda = 1 the results of z().

## Usage
Below is a basic demonstration of applying znorm transformations.

```{r setup, message = FALSE}
library(znorm)
```

We can transform laboratory measurements based on defined lower and upper reference limits:

```{r example}
# Example analyte measurements
vals <- c(10, 15, 20, 25, 30, 100)

# Calculate znorm values using Lower limit (L) and Upper limit (U)
znorm_scores <- znorm(vals, limits = c(12,15))

# Print standardized znorm scores
print(znorm_scores)
```
Use the same reference limits to recover the original measurements:

```{r inverse}
iznorm(znorm_scores, limits = c(12, 25))
```